Best Practices for Alchemical Free Energy Calculations [Article v \versionnumber]

### 2020-11-28 paper/manuscript.tex

\paragraph{Constraints and relative free energy calculations}
One issue which requires particular care is the use of constraints.
Commonly, bonds involving hydrogen are constrained to a fixed length using algorithms such as SHAKE or LINCS, allowing the use of longer timesteps~\cite{krautler2001fast}.
However, in single topology relative free energy calculations, the atoms involved might be mutated to other atom types -- for example, in a mutation of methane to methanol, one hydrogen might become an oxygen atom.
Some molecular dynamics engines are not set up to recognize this change, or at least not to correctly include contributions to the free energy from changing constraints/constraint length, so results for a transformation can be erroneous.
At present the most general solution to this problem is simply to avoid the use of constraints (and thus use a smaller timestep if necessary, usually of around 1 fs) in any relative free energy calculation involving a transformation of a constrained bond. Individual software programs and settings can handle such issues. For example, bonds can be transformed in both GROMACS and GROMOS, because contributions of LINCS (GROMACS) and SHAKE (GROMOS) contrained bonds are added to the $\frac{dH}{d\lambda}$ term~\cite{pearlman1993determining, straatsma1992holonomica, pearlman1991overlooked, gunsteren1997computer}. However, the constraints are not taken into account when calculating energy differences to other intermediate states (as the effect is entropic, not energetic). User manuals should carefully checked for how these effects are included if a constrained bond changes in length. We do note that if one performs calculations of the changing constraint in both solution and in the binding site, then the errors often cancel substantially, but the cancellation is not guaranteed.

>>>>

\paragraph{Constraints and relative free energy calculations}
One issue which requires particular care is the use of constraints.
Commonly, bonds involving hydrogen are constrained to a fixed length using algorithms such as SHAKE or LINCS, allowing the use of longer timesteps~\cite{krautler2001fast}.
However, in single topology relative free energy calculations, the atoms involved might be mutated to other atom types -- for example, in a mutation of methane to methanol, one hydrogen might become an oxygen atom.
Some molecular dynamics engines are not set up to recognize this change, or at least not to correctly include contributions to the free energy from changing constraints/constraint length, so results for a transformation can be erroneous.
At present the most general solution to this problem is simply to avoid the use of constraints (and thus use a smaller timestep if necessary, usually of around 1 fs) in any relative free energy calculation involving a transformation of a constrained bond. Individual software programs and settings can handle such issues. For example, bonds can be transformed in both GROMACS and GROMOS, because contributions of LINCS (GROMACS) and SHAKE (GROMOS) contrained bonds are added to the $\frac{dH}{d\lambda}$ term~\cite{pearlman1993determining, straatsma1992holonomica, pearlman1991overlooked, gunsteren1998computer}. However, the constraints are not taken into account when calculating energy differences to other intermediate states (as the effect is entropic, not energetic). User manuals should carefully checked for how these effects are included if a constrained bond changes in length. We do note that if one performs calculations of the changing constraint in both solution and in the binding site, then the errors often cancel substantially, but the cancellation is not guaranteed.
